---
title: "CoMMpass CGS high risk classification"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 6
    code_folding: hide
---

# Setup

```{r, warning = FALSE, message = FALSE, fig.height=8, fig.width=12}

library(tidyverse)
library(patchwork)
library(ComplexHeatmap)

proj_dir <- here::here()
processed_data_dir <- here::here(proj_dir, "processed_data")
plots_dir <- here::here(proj_dir, "plots")

input_filename <- here::here(processed_data_dir, "genome_cna_1q_1p32_17p13_clinical_df.tsv")
ancestry_filename <- here::here(processed_data_dir, "ancestry_df.tsv")
immune_atlas_samples_filename <- here::here(processed_data_dir, "immune_atlas_samples.tsv")

input_df <- read_tsv(input_filename)
ancestry_df <- read_tsv(ancestry_filename)
immune_atlas_samples_df <- read_tsv(immune_atlas_samples_filename)

```

# Creatinine

According to CGS, normal creatinine levels are < 1.2 mg/dL.
CoMMpass data reports creatinine as micro-mol/L.
The molar mass of creatinine is 113.12 g/mol, or 113.12e-6 g/micro-mol.
Therefore normal creatinine is < 1.2 mg/dL x (1g/1000mg) x (10 dL/L) x (1micro-mol/113.12e-6g) = 106.082 micro-mol/L
See violin plots:

```{r, warning = FALSE, message = FALSE, fig.height=8, fig.width=12}

set.seed(1)
creatinine_mgdl_plot <- input_df |>
  ggplot(aes(x = "",
             y = Creatinine)) +
  geom_violin() +
  geom_jitter(height = 0) +
  geom_hline(yintercept = 1.2,
             lty = 2,
             color = "red") +
  labs(x = NULL,
       y = "mg/dL",
       title = "Creatinine levels assuming mg/dL") +
  theme_bw()

set.seed(1)
creatinine_micromolL_plot <- input_df |>
  ggplot(aes(x = "",
             y = Creatinine)) +
  geom_violin() +
  geom_jitter(height = 0) +
  geom_hline(yintercept = 106.082,
             lty = 2,
             color = "red") +
  labs(x = NULL,
       y = "micromol/L",
       title = "Creatinine levels assuming micromol/L") +
  theme_bw()

creatinine_mgdl_plot + creatinine_micromolL_plot

```

# Evaluate risk criteria

- Criteria 1: del(17p), with a cutoff of >20% clonal fraction, and/or TP53 mutation
- Criteria 2: an IgH translocation including t(4;14), t(14;16), or t(14;20) along with 1q+ and/or del(1p32)
- Criteria 3: monoallelic del(1p32) along with 1q+ or biallelic del(1p32)
- Criteria 4: β2 microglobulin ≥5.5 mg/L with normal creatinine (<1.2 mg/dL)

Patients meeting one or more criteria are High Risk (HR).
Patients meeting no criteria and having no missing data are Standard Risk (SR).

```{r, warning = FALSE, message = FALSE, fig.height=8, fig.width=12}

risk_df <- input_df |>
  mutate(in_immune_atlas = public_id %in% immune_atlas_samples_df$public_id) |>
  mutate(has_del17p13 = ifelse(mean_log2_17p13 <= -0.152, TRUE, FALSE),
         has_TP53_mutation = case_when(TP53_Mutation_Count > 0 ~ TRUE,
                                       is.na(TP53_Mutation_Count) ~ NA,
                                       .default = FALSE),
         has_t414 = ifelse(NSD2_t_4_14_pos == 1, TRUE, FALSE),
         has_t1416 = ifelse(MAF_t_14_16_pos == 1, TRUE, FALSE),
         has_t1420 = ifelse(MAFB_t_14_20_pos == 1, TRUE, FALSE),
         has_1q = ifelse(mean_log2_1q >= 0.1375, TRUE, FALSE),
         has_mono_del1p32 = ifelse(mean_log2_1p32 <= -0.152, TRUE, FALSE),
         has_bi_del1p32 = ifelse(mean_log2_1p32 <= -1.3219, TRUE, FALSE),
         has_del1p32 = has_mono_del1p32 | has_bi_del1p32,
         has_high_B2M = ifelse(Serum_B2M >= 5.5, TRUE, FALSE),
         has_normal_creatinine = ifelse(Creatinine < 106.082, TRUE, FALSE),
         criteria1 = has_del17p13 | has_TP53_mutation,
         criteria2 = (has_t414 | has_t1416 | has_t1420) & (has_1q | has_del1p32),
         criteria3 = (has_mono_del1p32 & has_1q) | has_bi_del1p32,
         criteria4 = has_high_B2M & has_normal_creatinine) |>
  rowwise() |>
  mutate(CGS_risk = case_when(any(criteria1, criteria2, criteria3, criteria4, na.rm = TRUE) ~ "High",
                              all(!criteria1, !criteria2, !criteria3, !criteria4, na.rm = FALSE) ~ "Standard",
                              .default = NA),
         CGS_risk_genomic = case_when(any(criteria1, criteria2, criteria3, na.rm = TRUE) ~ "High",
                                      all(!criteria1, !criteria2, !criteria3, na.rm = FALSE) ~ "Standard",
                                      .default = NA),
         CGS_risk_C4 = case_when(criteria4 & all(!criteria1, !criteria2, !criteria3, na.rm = FALSE) ~ "High",
                                 all(!criteria1, !criteria2, !criteria3, !criteria4, na.rm = FALSE) ~ "Standard",
                                 .default = NA)) |>
  ungroup() |>
  left_join(ancestry_df |> select(public_id, afr_high),
            by = c("public_id"))

readr::write_tsv(x = risk_df,
                 file = here::here(processed_data_dir, "CGS_risk_full_criteria_df.tsv"))

readr::write_tsv(x = risk_df |> select(public_id, CGS_risk, CGS_risk_genomic),
                 file = here::here(processed_data_dir, "CGS_risk_df.tsv"))

readr::write_tsv(x = risk_df |> select(public_id, 
                                       in_immune_atlas, 
                                       afr_high, 
                                       CGS_risk, 
                                       starts_with("criteria"),
                                       starts_with("has")),
                 file = here::here(processed_data_dir, "CGS_risk_key_criteria.tsv"))

```

# Ancestry concordance

```{r, warning = FALSE, message = FALSE}

genetic_ancestry_concordant_with_self_reported_ethnicity_prop <- risk_df |> 
  filter(reclassify_ethnicity %in% c("Black", "White", "Asian")) |> 
  #count(afr_high, reclassify_ethnicity) |>
  mutate(concordant = case_when(afr_high == "AFR High" & reclassify_ethnicity %in% c("Black") ~ TRUE,
                                afr_high == "AFR Low" & reclassify_ethnicity %in% c("White", "Asian") ~ TRUE,
                                .default = FALSE)) |>
  pull(concordant) |>
  mean()

additional_samples_with_genetic_ancestry <- risk_df |> 
  count(afr_high, reclassify_ethnicity) |> 
  filter(is.na(reclassify_ethnicity) | 
           reclassify_ethnicity %in% c("Other_Unknown")) |> 
  filter(!is.na(afr_high)) |> 
  pull(n) |> 
  sum()

str_c("Genetic ancestry concordance with self-reported ethnicity: ",
      round(100*genetic_ancestry_concordant_with_self_reported_ethnicity_prop, 2), "%. ",
      "Additional samples with genetic ancestry: ",
      additional_samples_with_genetic_ancestry)

```

# Explore a bit

```{r, warning = FALSE, message = FALSE, fig.height=8, fig.width=12}

gender_plot <- risk_df |>
  count(CGS_risk, Gender) |>
  ggplot(aes(x = Gender,
             y = n,
             fill = CGS_risk)) +
  geom_col(position = "fill") +
  labs(title = "Gender",
       x = "Gender",
       y = "Proportion",
       fill = "CGS risk") +
  theme_bw()

ethnicity_plot <- risk_df |>
  count(CGS_risk, reclassify_ethnicity) |>
  ggplot(aes(x = reclassify_ethnicity,
             y = n,
             fill = CGS_risk)) +
  geom_col(position = "fill") +
  labs(title = "Ethnicity",
       x = "Ethnicity",
       y = "Proportion",
       fill = "CGS risk") +
  theme_bw()

iss_stage_plot <- risk_df |>
  count(CGS_risk, ISS_Stage) |>
  ggplot(aes(x = factor(ISS_Stage),
             y = n,
             fill = CGS_risk)) +
  geom_col(position = "fill") +
  labs(title = "ISS stage",
       x = "ISS stage",
       y = "Proportion",
       fill = "CGS risk") +
  theme_bw()

imwg_risk_plot <- risk_df |>
  count(CGS_risk, IMWG_Risk_Class) |>
  ggplot(aes(x = factor(IMWG_Risk_Class),
             y = n,
             fill = CGS_risk)) +
  geom_col(position = "fill") +
  labs(title = "IMWG risk (former definition)",
       x = "IMWG risk\n(former definition)",
       y = "Proportion",
       fill = "CGS risk") +
  theme_bw()

age_plot <- risk_df |>
  ggplot(aes(x = CGS_risk,
             y = Age,
             color = CGS_risk)) +
  geom_violin(draw_quantiles = 0.5,
              show.legend = FALSE) +
  geom_jitter(height = 0,
              width = 0.1,
              show.legend = FALSE) +
  labs(title = "Age",
       x = "CGS risk") +
  theme_bw()

cytogenetic_risk_plot <- risk_df |>
  count(CGS_risk, Cytogenetic_High_Risk) |>
  ggplot(aes(x = factor(Cytogenetic_High_Risk),
             y = n,
             fill = CGS_risk)) +
  geom_col(position = "fill") +
  labs(title = "Cytogenetic risk",
       x = "Cytogenetic risk",
       fill = "CGS risk",
       y = "Proportion") +
  theme_bw()

gender_plot + ethnicity_plot + iss_stage_plot + 
  imwg_risk_plot + cytogenetic_risk_plot + age_plot +
  plot_layout(guides = "collect") +
  plot_annotation(title = "CGS risk (including all criteria)")

```

# AFR plot

```{r, warning = FALSE, message = FALSE, fig.height=8, fig.width=12}

prop_text_genomic_df <- risk_df |>
  filter(!is.na(CGS_risk_genomic), !is.na(afr_high)) |>
  rbind(risk_df |>
          filter(!is.na(CGS_risk_genomic), !is.na(afr_high)) |>
          mutate(afr_high = "Overall")) |>
  count(CGS_risk_genomic, afr_high) |>
  group_by(afr_high) |>
  mutate(prop = n/sum(n))

flip_afr_risk_genomic_plot <- risk_df |>
  filter(!is.na(CGS_risk_genomic), !is.na(afr_high)) |>
  rbind(risk_df |>
          filter(!is.na(CGS_risk_genomic), !is.na(afr_high)) |>
          mutate(afr_high = "Overall")) |>
  count(CGS_risk_genomic, afr_high) |>
  ggplot(aes(fill = CGS_risk_genomic,
             y = n,
             x = afr_high)) +
  geom_col(position = "fill") +
  geom_text(data = prop_text_genomic_df |> filter(CGS_risk_genomic == "High"),
            aes(x = afr_high,
                y = 1 - prop + .03,
                label = str_c(round(100*prop, 1), "%, n = ", n)),
            vjust = 0) +
  geom_text(data = prop_text_genomic_df |> filter(CGS_risk_genomic == "Standard"),
            aes(x = afr_high,
                y = prop - .03,
                label = str_c(round(100*prop, 1), "%, n = ", n)),
            vjust = 1) +
  scale_y_continuous(breaks = seq(0,1,0.1)) +
  labs(title = "CGS HR vs. SR within AFR groups",
       subtitle = "Genomic only",
       fill = "CGS risk",
       x = "1KG-AFR-like",
       y = "Proportion") +
  theme_bw()

prop_text_df <- risk_df |>
  filter(!is.na(CGS_risk), !is.na(afr_high)) |>
  rbind(risk_df |>
          filter(!is.na(CGS_risk), !is.na(afr_high)) |>
          mutate(afr_high = "Overall")) |>
  count(CGS_risk, afr_high) |>
  group_by(afr_high) |>
  mutate(prop = n/sum(n))

flip_afr_risk_plot <- risk_df |>
  filter(!is.na(CGS_risk), !is.na(afr_high)) |>
  rbind(risk_df |>
          filter(!is.na(CGS_risk), !is.na(afr_high)) |>
          mutate(afr_high = "Overall")) |>
  count(CGS_risk, afr_high) |>
  ggplot(aes(fill = CGS_risk,
             y = n,
             x = afr_high)) +
  geom_col(position = "fill") +
  geom_text(data = prop_text_df |> filter(CGS_risk == "High"),
            aes(x = afr_high,
                y = 1 - prop + .03,
                label = str_c(round(100*prop, 1), "%, n = ", n)),
            vjust = 0) +
  geom_text(data = prop_text_df |> filter(CGS_risk == "Standard"),
            aes(x = afr_high,
                y = prop - .03,
                label = str_c(round(100*prop, 1), "%, n = ", n)),
            vjust = 1) +
  scale_y_continuous(breaks = seq(0,1,0.1)) +
  labs(title = "CGS HR vs. SR within AFR groups",
       subtitle = "All criteria",
       fill = "CGS risk",
       x = "1KG-AFR-like",
       y = "Proportion") +
  theme_bw()

ggsave(filename = here::here(plots_dir, "flip_afr_risk.pdf"),
       plot = flip_afr_risk_plot + flip_afr_risk_genomic_plot + plot_layout(guides = "collect"),
       width = 12, height = 6)

flip_afr_risk_plot + flip_afr_risk_genomic_plot + plot_layout(guides = "collect")

```

# Criteria by ancestry

```{r, warning = FALSE, message = FALSE}

events_of_interest <- c("criteria1", "has_del17p13", "has_TP53_mutation",
                        "criteria2", "has_t414", "has_t1416", "has_t1420", "has_1q", "has_del1p32",
                        "criteria3", "has_mono_del1p32", #"has_bi_del1p32",
                        "criteria4", "has_high_B2M", "has_normal_creatinine")

events_of_interest |>
  purrr::map(
    \(event_of_interest) {
      
      event_table <- risk_df |>
        filter(!is.na(afr_high),
               !is.na(!!sym(event_of_interest))) |>
        count(afr_high,
              !!sym(event_of_interest)) |>
        group_by(afr_high) |>
        mutate(total = sum(n),
               prop = n/sum(n),
               hr_factor = ifelse(!!sym(event_of_interest), "HR factor (TRUE)", "SR factor (FALSE)"))
      
      p_value <- fisher.test(matrix(event_table$n, nrow = 2))$p.value
      
      event_table |> 
        ggplot(aes(fill = hr_factor,
                   y = n,
                   x = afr_high)) +
        geom_col(position = "fill") +
        geom_text(data = event_table |> filter(!!sym(event_of_interest) == FALSE),
                  aes(x = afr_high,
                      y = prop - .03,
                      label = str_c(round(100*prop, 1), "%, n = ", n)),
                  vjust = 1) +
        geom_text(data = event_table |> filter(!!sym(event_of_interest) == TRUE),
                  aes(x = afr_high,
                      y = 1 - prop + .03,
                      label = str_c(round(100*prop, 1), "%, n = ", n)),
                  vjust = 0) +
        scale_y_continuous(breaks = seq(0,1,0.1)) +
        #scale_fill_manual(values = c("#F8766D", "#00BFC4")) +
        labs(title = event_of_interest,
             subtitle = str_c("p-value = ", round(p_value, 3)),
             fill = event_of_interest,
             x = "1KG-AFR-like",
             y = "Proportion") +
        theme_bw()
      
      
    } ) |> 
  setNames(events_of_interest)



```

#### Creatinine by ancestry and B2M

```{r, warning = FALSE, message = FALSE}

# COMMENT FROM GEORGE: Off topic - for later: in the Afr-high with high beta 2 AND also high creatinine, what was PFS/OS compared to SR and compared to high beta-2 and norm creatinine?  It could be the high creatinine cutoff (gneerally determined based on white patients) is not meaningful and these patients have HR clinical outcomes. (If they do - we would have to make sure they still got good dosage of drugs - rather than worse outcomes because of under-treatment - docs worried about kidney failure...)

set.seed(1)

risk_df |> 
  filter(!is.na(afr_high), !is.na(has_high_B2M), !is.na(has_normal_creatinine), !is.na(Creatinine)) |>
  mutate(has_high_B2M = ifelse(has_high_B2M, "High B2M", "Normal B2M")) |> 
  ggplot(aes(x = afr_high, 
             y = log(Creatinine))) + 
  geom_violin(draw_quantiles = 0.5) + 
  geom_point(aes(color = criteria4), 
             #position = position_jitterdodge(dodge.width = 0.9),
             position = position_jitter(height = 0, width = 0.1), 
             shape = 16, 
             alpha = 0.75, 
             show.legend = TRUE) + 
  geom_hline(yintercept = log(106.082), 
             lty = 2, 
             color = "red") + 
  facet_wrap(~ has_high_B2M) + 
  labs(title = "Creatinine levels",
       subtitle = "by genetic ancestry and B2M levels",
       x = "Genetic ancestry", 
       y = "log(Creatinine) (micro-mol/L)") + 
  scale_color_brewer(type = "qual") + 
  theme_bw(base_size = 20)


```

# Creatinine for publication

```{r}

ancestry_colors <- c("AFR High" = "#86325F", "AFR Low" = "#5A90C7")

creatinine_plot <- risk_df |>
  mutate(cohort = "Full CoMMpass cohort") |>
  bind_rows(risk_df |>
              filter(public_id %in% immune_atlas_samples_df$public_id) |>
              mutate(cohort = "Immune Atlas subset")) |>
  filter(!is.na(afr_high), !is.na(has_high_B2M), !is.na(has_normal_creatinine), !is.na(Creatinine)) |>
  mutate(has_high_B2M = ifelse(has_high_B2M, "High B2M", "Normal B2M")) |>
  ggplot(aes(fill = afr_high,
             x = has_high_B2M,
             y = Creatinine)) +
  geom_violin(draw_quantiles = 0.5,
              scale = "width",
              alpha = 0.5) +
  geom_hline(yintercept = 106.082,
             lty = 2,
             color = "red") +
  geom_point(position = position_jitterdodge(dodge.width = 0.9, jitter.width = 0.1, jitter.height = 0, seed = 1),
             shape = 16,
             show.legend = FALSE) +
  scale_y_continuous(limits = c(0, 2100)) +
  ggbreak::scale_y_break(c(1000, 1000), scales = 0.25, ticklabels = c(1250, 1500, 1750, 2000)) +
  annotate(geom = "text", x = 0, y = 106.082*1.2, label = "   High", hjust = 0, vjust = 0) +
  annotate(geom = "text", x = 0, y = 106.082*.8, label = "   Normal", hjust = 0, vjust = 1) +
  facet_wrap(~ cohort) +
  labs(title = "Criteria 4: High B2M with normal creatinine",
       x = "Genetic ancestry",
       y = "Creatinine (micro-mol/L)",
       fill = "Ancestry") +
  scale_fill_manual(values = ancestry_colors) +
  theme_bw(base_size = 12)

ggsave(plot = creatinine_plot,
       filename = here::here(plots_dir, "creatinine.pdf"), 
       width = 10, height = 6)

creatinine_plot

```

# Test for difference in ancestry

#### All criteria

```{r, warning = FALSE, message = FALSE, fig.height=8, fig.width=12}

afr_high_CGS_table <- risk_df |>
  select(afr_high, CGS_risk) |>
  table()

list(afr_high_CGS_table, 
     fisher.test(afr_high_CGS_table), 
     chisq.test(afr_high_CGS_table))

```

#### Genomics only

```{r, warning = FALSE, message = FALSE, fig.height=8, fig.width=12}

afr_high_CGS_genomic_table <- risk_df |>
  select(afr_high, CGS_risk_genomic) |>
  table()

list(afr_high_CGS_genomic_table,
     fisher.test(afr_high_CGS_genomic_table),
     chisq.test(afr_high_CGS_genomic_table))

```

# Criteria overlap 

```{r, warning = FALSE, message = FALSE, fig.height=8, fig.width=12}

# UpSet plot
m <- make_comb_mat(risk_df |>
                     filter(!is.na(CGS_risk)) |>
                     select(criteria1,
                            criteria2,
                            criteria3,
                            criteria4))

UpSet(m = m,
      top_annotation = upset_top_annotation(m, add_numbers = TRUE, ),
      right_annotation = upset_right_annotation(m, add_numbers = TRUE))

```

# Heatmap

```{r, warning = FALSE, message = FALSE, fig.height=8, fig.width=12}

heatmap_df <- risk_df |>
  select(public_id, starts_with("criteria")) |>
  column_to_rownames(var = "public_id") |>
  mutate(across(.cols = everything(), .fns = as.numeric)) |>
  t()

col_annotation_df <- risk_df |>
  select(public_id, 
         has_del17p13,
         has_TP53_mutation,
         has_t414,
         has_t1416,
         has_t1420,
         has_1q,
         has_del1p32,
         has_mono_del1p32,
         has_bi_del1p32,
         has_high_B2M,
         has_normal_creatinine,
         afr_high, 
         CGS_risk) |>
  column_to_rownames(var = "public_id") |>
  mutate(across(.cols = everything(), .fns = factor))

heatmap_plot <- pheatmap::pheatmap(
  heatmap_df,
  main = "CoMMpass (n = 877): CGS High Risk Multiple Myeloma Definition",
  labels_row = c("Criteria 1:\ndel(17p) or TP53 mutation",
                 "Criteria 2:\nt(4;14), t(14;16), or t(14;20)\nalong with 1q+ and/or del(1p32)",
                 "Criteria 3:\nmonoallelic del(1p32)\nalong with 1q+ or biallelic del(1p32)",
                 "Criteria 4:\nB2M ≥5.5 mg/L\nwith normal creatinine (<1.2 mg/dL)"),
  annotation_col = col_annotation_df,
  fontsize = 16,
  legend = FALSE,
  annotation_legend = FALSE,
  legend_breaks = c(0,1),
  show_colnames = FALSE,
  color = c("#bcbddc", "#756bb1"),
  annotation_colors = list("CGS_risk" = c("High" = "#de2d26", "Standard" = "#fee0d2"),
                           "afr_high" = c("AFR High" = "black", "AFR Low" = "grey90"),
                           "has_normal_creatinine" = c("TRUE" = "#31a354", "FALSE" = "#a1d99b"),
                           "has_high_B2M" = c("TRUE" = "#31a354", "FALSE" = "#a1d99b"),
                           "has_bi_del1p32" = c("TRUE" = "#2c7fb8", "FALSE" = "#edf8b1"),
                           "has_mono_del1p32" = c("TRUE" = "#2c7fb8", "FALSE" = "#edf8b1"),
                           "has_del1p32" = c("TRUE" = "#e6550d", "FALSE" = "#fee6ce"),
                           "has_1q" = c("TRUE" = "#e6550d", "FALSE" = "#fee6ce"),
                           "has_t1416" = c("TRUE" = "#e6550d", "FALSE" = "#fee6ce"),
                           "has_t1420" = c("TRUE" = "#e6550d", "FALSE" = "#fee6ce"),
                           "has_t414" = c("TRUE" = "#e6550d", "FALSE" = "#fee6ce"),
                           "has_TP53_mutation" = c("TRUE" = "#3182bd", "FALSE" = "#deebf7"),
                           "has_del17p13" = c("TRUE" = "#3182bd", "FALSE" = "#deebf7")))

ggsave(filename = here::here(plots_dir, "cgs_heatmap.pdf"), 
       plot = heatmap_plot, 
       height = 8, 
       width = 12)

heatmap_plot

```

# Survival

```{r, warning = FALSE, message = FALSE, fig.height=8, fig.width=12}

# survival (PFS)

Surv_obj <- with(risk_df,
                 survival::Surv(final_survival_time_pfs, final_censor_pfs))

survfit_obj <- survminer::surv_fit(Surv_obj ~ CGS_risk,
                                   data = risk_df)

abstract_PFS_object <- survfit_obj

pfs_plot <- survminer::ggsurvplot(
  fit = survfit_obj,
  data = risk_df,
  conf.int = TRUE,
  pval = TRUE,
  surv.median.line = "hv",
  ggtheme = theme_bw(),
  risk.table = TRUE,
  title = "Progression-free Survival (PFS)",
  xlab = "Time (days)",
  ylab = "Probability",
  legend.title = "")

# survival (OS)

Surv_obj <- with(risk_df,
                 survival::Surv(final_survival_time_os, final_censor_os))

survfit_obj <- survminer::surv_fit(Surv_obj ~ CGS_risk,
                                   data = risk_df)

abstract_OS_object <- survfit_obj

os_plot <- survminer::ggsurvplot(
  fit = survfit_obj,
  data = risk_df,
  conf.int = TRUE,
  pval = TRUE,
  surv.median.line = "hv",
  ggtheme = theme_bw(),
  risk.table = TRUE,
  title = "Overall Survival (OS)",
  xlab = "Time (days)",
  ylab = "Probability",
  legend.title = "")

combined_survival_plot_object <- (pfs_plot$plot + os_plot$plot) / free(pfs_plot$table + os_plot$table) +
  plot_layout(guides = "collect", heights = c(4,2)) +
  plot_annotation(title = "CGS high risk vs. standard risk") &
  theme_bw(base_size = 16) +
  theme(legend.position = "top")

ggsave(filename = here::here(plots_dir, str_c("CGS_HR_SR_PFS_OS.pdf")),
       plot = combined_survival_plot_object,
       width = 12,
       height = 8)

combined_survival_plot_object

```

# Survival triplet ASCT
```{r, warning = FALSE, message = FALSE, fig.height=8, fig.width=12}

# survival (PFS)

Surv_obj <- with(risk_df |> filter(received_asct, received_triplet),
                 survival::Surv(final_survival_time_pfs, final_censor_pfs))

survfit_obj <- survminer::surv_fit(Surv_obj ~ CGS_risk,
                                   data = risk_df |> filter(received_asct, received_triplet))

summary(survfit_obj)$table[,"median"]/30.4375

pfs_plot <- survminer::ggsurvplot(
  fit = survfit_obj,
  data = risk_df |> filter(received_asct, received_triplet),
  conf.int = TRUE,
  pval = TRUE,
  surv.median.line = "hv",
  ggtheme = theme_bw(),
  risk.table = TRUE,
  title = "Progression-free Survival (PFS)",
  xlab = "Time (days)",
  ylab = "Probability",
  legend.title = "")

# survival (OS)

Surv_obj <- with(risk_df |> filter(received_asct, received_triplet),
                 survival::Surv(final_survival_time_os, final_censor_os))

survfit_obj <- survminer::surv_fit(Surv_obj ~ CGS_risk,
                                   data = risk_df |> filter(received_asct, received_triplet))

summary(survfit_obj)$table[,"median"]/30.4375

os_plot <- survminer::ggsurvplot(
  fit = survfit_obj,
  data = risk_df |> filter(received_asct, received_triplet),
  conf.int = TRUE,
  pval = TRUE,
  surv.median.line = "hv",
  ggtheme = theme_bw(),
  risk.table = TRUE,
  title = "Overall Survival (OS)",
  xlab = "Time (days)",
  ylab = "Probability",
  legend.title = "")

combined_survival_plot_object <- (pfs_plot$plot + os_plot$plot) / free(pfs_plot$table + os_plot$table) +
  plot_layout(guides = "collect", heights = c(4,2)) +
  plot_annotation(title = "CGS high risk vs. standard risk") &
  theme_bw(base_size = 16) +
  theme(legend.position = "top")

ggsave(filename = here::here(plots_dir, str_c("CGS_HR_SR_PFS_OS.ASCT_triplet_only.pdf")),
       plot = combined_survival_plot_object,
       width = 12,
       height = 8)

combined_survival_plot_object

```

# Survival by AFR

```{r, warning = FALSE, message = FALSE, fig.height=8, fig.width=12}

# survival (PFS) by AFR

Surv_obj <- with(risk_df,
                 survival::Surv(final_survival_time_pfs, final_censor_pfs))

survfit_obj <- survminer::surv_fit(Surv_obj ~ CGS_risk + afr_high,
                                   data = risk_df)

pfs_plot <- survminer::ggsurvplot(
  fit = survfit_obj,
  data = risk_df,
  conf.int = TRUE,
  pval = TRUE,
  surv.median.line = "hv",
  ggtheme = theme_bw(),
  risk.table = TRUE,
  title = "Progression-free Survival (PFS)",
  xlab = "Time (days)",
  ylab = "Probability",
  legend.title = "")

# survival (OS) by AFR

Surv_obj <- with(risk_df,
                 survival::Surv(final_survival_time_os, final_censor_os))

survfit_obj <- survminer::surv_fit(Surv_obj ~ CGS_risk + afr_high,
                                   data = risk_df)

os_plot <- survminer::ggsurvplot(
  fit = survfit_obj,
  data = risk_df,
  conf.int = TRUE,
  pval = TRUE,
  surv.median.line = "hv",
  ggtheme = theme_bw(),
  risk.table = TRUE,
  title = "Overall Survival (OS)",
  xlab = "Time (days)",
  ylab = "Probability",
  legend.title = "")

combined_survival_plot_object <- (pfs_plot$plot + os_plot$plot) / free(pfs_plot$table + os_plot$table) +
  plot_layout(guides = "collect", heights = c(4,2)) +
  plot_annotation(title = "CGS high risk vs. standard risk", subtitle = "by AFR High status") &
  theme_bw(base_size = 16) +
  theme(legend.position = "top", legend.direction = "vertical")

ggsave(filename = here::here(plots_dir, str_c("CGS_HR_SR_PFS_OS.by_AFR.pdf")),
       plot = combined_survival_plot_object,
       width = 16,
       height = 12)

combined_survival_plot_object

```

# Survival by AFR (SR only)

```{r, warning = FALSE, message = FALSE, fig.height=8, fig.width=12}

# survival (PFS) by AFR (standard risk only)

Surv_obj <- with(risk_df |> filter(CGS_risk == "Standard"),
                 survival::Surv(final_survival_time_pfs, final_censor_pfs))

survfit_obj <- survminer::surv_fit(Surv_obj ~ CGS_risk + afr_high,
                                   data = risk_df |> filter(CGS_risk == "Standard"))

pfs_plot <- survminer::ggsurvplot(
  fit = survfit_obj,
  data = risk_df |> filter(CGS_risk == "Standard"),
  conf.int = TRUE,
  pval = TRUE,
  surv.median.line = "hv",
  ggtheme = theme_bw(),
  risk.table = TRUE,
  title = "Progression-free Survival (PFS)",
  xlab = "Time (days)",
  ylab = "Probability",
  legend.title = "")

# survival (OS) by AFR (standard risk only)

Surv_obj <- with(risk_df |> filter(CGS_risk == "Standard"),
                 survival::Surv(final_survival_time_os, final_censor_os))

survfit_obj <- survminer::surv_fit(Surv_obj ~ CGS_risk + afr_high,
                                   data = risk_df |> filter(CGS_risk == "Standard"))

os_plot <- survminer::ggsurvplot(
  fit = survfit_obj,
  data = risk_df |> filter(CGS_risk == "Standard"),
  conf.int = TRUE,
  pval = TRUE,
  surv.median.line = "hv",
  ggtheme = theme_bw(),
  risk.table = TRUE,
  title = "Overall Survival (OS)",
  xlab = "Time (days)",
  ylab = "Probability",
  legend.title = "")

combined_survival_plot_object <- (pfs_plot$plot + os_plot$plot) / free(pfs_plot$table + os_plot$table) +
  plot_layout(guides = "collect", heights = c(4,2)) +
  plot_annotation(title = "CGS Standard risk", subtitle = "by AFR High status") &
  theme_bw(base_size = 16) +
  theme(legend.position = "top", legend.direction = "vertical")

ggsave(filename = here::here(plots_dir, str_c("CGS_SR_only_PFS_OS.by_AFR.pdf")),
       plot = combined_survival_plot_object,
       width = 16,
       height = 12)

combined_survival_plot_object

```

# Survival by AFR (HR only)

```{r, warning = FALSE, message = FALSE, fig.height=8, fig.width=12}

# survival (PFS) by AFR (high risk only)

Surv_obj <- with(risk_df |> filter(CGS_risk == "High"),
                 survival::Surv(final_survival_time_pfs, final_censor_pfs))

survfit_obj <- survminer::surv_fit(Surv_obj ~ CGS_risk + afr_high,
                                   data = risk_df |> filter(CGS_risk == "High"))

pfs_plot <- survminer::ggsurvplot(
  fit = survfit_obj,
  data = risk_df |> filter(CGS_risk == "High"),
  conf.int = TRUE,
  pval = TRUE,
  surv.median.line = "hv",
  ggtheme = theme_bw(),
  risk.table = TRUE,
  title = "Progression-free Survival (PFS)",
  xlab = "Time (days)",
  ylab = "Probability",
  legend.title = "")

# survival (OS) by AFR (high risk only)

Surv_obj <- with(risk_df |> filter(CGS_risk == "High"),
                 survival::Surv(final_survival_time_os, final_censor_os))

survfit_obj <- survminer::surv_fit(Surv_obj ~ CGS_risk + afr_high,
                                   data = risk_df |> filter(CGS_risk == "High"))

os_plot <- survminer::ggsurvplot(
  fit = survfit_obj,
  data = risk_df |> filter(CGS_risk == "High"),
  conf.int = TRUE,
  pval = TRUE,
  surv.median.line = "hv",
  ggtheme = theme_bw(),
  risk.table = TRUE,
  title = "Overall Survival (OS)",
  xlab = "Time (days)",
  ylab = "Probability",
  legend.title = "")

combined_survival_plot_object <- (pfs_plot$plot + os_plot$plot) / free(pfs_plot$table + os_plot$table) +
  plot_layout(guides = "collect", heights = c(4,2)) +
  plot_annotation(title = "CGS High risk", subtitle = "by AFR High status") &
  theme_bw(base_size = 16) +
  theme(legend.position = "top", legend.direction = "vertical")

ggsave(filename = here::here(plots_dir, str_c("CGS_HR_only_PFS_OS.by_AFR.pdf")),
       plot = combined_survival_plot_object,
       width = 16,
       height = 12)

combined_survival_plot_object

```

# Cox PH model

Does AFR ancestry impact survival controlling for Age, Transplant, and Triplet therapy?

#### SR and HR PFS
```{r, warning = FALSE, message = FALSE, fig.height=8, fig.width=12}

Surv_obj <- with(risk_df,
                 survival::Surv(final_survival_time_pfs, final_censor_pfs))

coxph_result <- survival::coxph(Surv_obj ~ Age + Gender + afr_high + CGS_risk + received_asct + received_triplet,
                                data = risk_df)

print(summary(coxph_result))

```

#### SR and HR OS
```{r, warning = FALSE, message = FALSE, fig.height=8, fig.width=12}

Surv_obj <- with(risk_df,
                 survival::Surv(final_survival_time_os, final_censor_os))

coxph_result <- survival::coxph(Surv_obj ~ Age + Gender + afr_high + CGS_risk + received_asct + received_triplet,
                                data = risk_df)

print(summary(coxph_result))

```


#### HR PFS

```{r, warning = FALSE, message = FALSE, fig.height=8, fig.width=12}

Surv_obj <- with(risk_df |> filter(CGS_risk == "High"),
                 survival::Surv(final_survival_time_pfs, final_censor_pfs))

coxph_result <- survival::coxph(Surv_obj ~ Age + Gender + afr_high + received_asct + received_triplet,
                                data = risk_df |> filter(CGS_risk == "High"))

print(summary(coxph_result))

```

#### SR PFS

```{r, warning = FALSE, message = FALSE, fig.height=8, fig.width=12}

Surv_obj <- with(risk_df |> filter(CGS_risk == "Standard"),
                 survival::Surv(final_survival_time_os, final_censor_os))

coxph_result <- survival::coxph(Surv_obj ~ Age + Gender + afr_high + received_asct + received_triplet,
                                data = risk_df |> filter(CGS_risk == "Standard"))

print(summary(coxph_result))

```

# Survival by criteria

```{r, warning = FALSE, message = FALSE, fig.height=8, fig.width=12}

criteria_list <- list("criteria1" = "Criteria 1: del(17p) or TP53 mutation",
                      "criteria2" = "Criteria 2: t(4;14), t(14;16), or t(14;20) along with 1q+ and/or del(1p32)",
                      "criteria3" = "Criteria 3: monoallelic del(1p32) along with 1q+ or biallelic del(1p32)",
                      "criteria4" = "Criteria 4: B2M ≥5.5 mg/L with normal creatinine (<1.2 mg/dL)")

criteria_survival_plot_list <- names(criteria_list) |>
  purrr::map(\(x) {
    
    renamed_criteria <- str_replace(x, "criteria", "Criteria ")
    
    criteria_df <- risk_df |>
      filter(get(x) | CGS_risk == "Standard") |>
      select(final_survival_time_pfs, final_censor_pfs, final_survival_time_os, final_censor_os, CGS_risk) |>
      mutate(final_survival_time_pfs = final_survival_time_pfs,
             final_survival_time_os = final_survival_time_os,
             CGS_risk = ifelse(CGS_risk == "High",
                               renamed_criteria,
                               "Standard"))
    
    pfs_Surv_obj <- with(criteria_df,
                         survival::Surv(final_survival_time_pfs, final_censor_pfs))
    
    pfs_survfit_obj <- survminer::surv_fit(pfs_Surv_obj ~ CGS_risk,
                                           data = criteria_df)
    
    pfs_plot <- survminer::ggsurvplot(
      fit = pfs_survfit_obj,
      data = criteria_df,
      conf.int = TRUE,
      pval = TRUE,
      surv.median.line = "hv",
      ggtheme = theme_bw(base_size = 16),
      risk.table = TRUE,
      title = "Progression-free Survival (PFS)",
      xlab = "Time (days)",
      ylab = "Probability (PFS)",
      legend.title = "")
    
    os_Surv_obj <- with(criteria_df,
                        survival::Surv(final_survival_time_os, final_censor_os))
    
    os_survfit_obj <- survminer::surv_fit(os_Surv_obj ~ CGS_risk,
                                          data = criteria_df)
    
    os_plot <- survminer::ggsurvplot(
      fit = os_survfit_obj,
      data = criteria_df,
      conf.int = TRUE,
      pval = TRUE,
      surv.median.line = "hv",
      ggtheme = theme_bw(base_size = 16),
      risk.table = TRUE,
      title = "Overall Survival (OS)",
      xlab = "Time (days)",
      ylab = "Probability (OS)",
      legend.title = "")
    
    combined_survival_plot_object <- (pfs_plot$plot + os_plot$plot) / free(pfs_plot$table + os_plot$table) +
      plot_layout(guides = "collect", heights = c(4,2)) +
      plot_annotation(title = str_c(renamed_criteria, " HR vs. SR"),
                      subtitle = criteria_list[[x]]) &
      theme_bw(base_size = 16) +
      theme(legend.position = "top")
    
    ggsave(filename = here::here(plots_dir, str_c(x, "_HR_SR_PFS_OS.pdf")),
           plot = combined_survival_plot_object,
           width = 12,
           height = 8)
    
    print(combined_survival_plot_object)
    
  }
  
  
  ) |>
  setNames(names(criteria_list))


```

Unfortunately, sample size is too low to look at AFR High vs. Low within each criteria.

# Number of criteria

```{r, warning = FALSE, message = FALSE, fig.height=8, fig.width=12}

criteria_df <- risk_df |>
  filter(!is.na(CGS_risk)) |>
  select(public_id,
         starts_with("criteria"),
         CGS_risk) |>
  pivot_longer(cols = starts_with("criteria")) |>
  group_by(public_id,
           CGS_risk) |>
  summarize(total = sum(value, na.rm = TRUE),
            total_na = sum(is.na(value))) |>
  ungroup() |>
  left_join(risk_df |> select(public_id, final_survival_time_pfs, final_censor_pfs, final_survival_time_os, final_censor_os),
            by = "public_id") |>
  filter(total %in% c(0,1,2))

pfs_Surv_obj <- with(criteria_df,
                     survival::Surv(final_survival_time_pfs/30.5, final_censor_pfs))

pfs_survfit_obj <- survminer::surv_fit(pfs_Surv_obj ~ total,
                                       data = criteria_df)

pfs_plot <- survminer::ggsurvplot(
  fit = pfs_survfit_obj,
  data = criteria_df,
  conf.int = TRUE,
  pval = TRUE,
  surv.median.line = "hv",
  ggtheme = theme_bw(base_size = 20),
  risk.table = TRUE,
  title = "Progression-free Survival (PFS)",
  xlab = "Time (days)",
  ylab = "Probability (PFS)",
  legend.title = "")

os_Surv_obj <- with(criteria_df,
                    survival::Surv(final_survival_time_pfs/30.5, final_censor_os))

os_survfit_obj <- survminer::surv_fit(os_Surv_obj ~ total,
                                      data = criteria_df)

os_plot <- survminer::ggsurvplot(
  fit = os_survfit_obj,
  data = criteria_df,
  conf.int = TRUE,
  pval = TRUE,
  surv.median.line = "hv",
  ggtheme = theme_bw(base_size = 20),
  risk.table = TRUE,
  title = "Overall Survival (PFS)",
  xlab = "Time (days)",
  ylab = "Probability (PFS)",
  legend.title = "")

(pfs_plot$plot + os_plot$plot) / free(pfs_plot$table + os_plot$table) +
  plot_layout(guides = "collect", heights = c(4,2)) +
  plot_annotation(title = "Is survival worse when more HR criteria are met?") &
  theme_bw(base_size = 16) +
  theme(legend.position = "top")

```

# C4 only vs. SR

```{r, warning = FALSE, message = FALSE, fig.height=8, fig.width=12}

criteria_df <- risk_df

pfs_Surv_obj <- with(criteria_df,
                     survival::Surv(final_survival_time_pfs, final_censor_pfs))

pfs_survfit_obj <- survminer::surv_fit(pfs_Surv_obj ~ CGS_risk_C4,
                                       data = criteria_df)

pfs_plot <- survminer::ggsurvplot(
  fit = pfs_survfit_obj,
  data = criteria_df,
  conf.int = TRUE,
  pval = TRUE,
  surv.median.line = "hv",
  ggtheme = theme_bw(base_size = 16),
  risk.table = TRUE,
  title = "Progression-free Survival (PFS)",
  xlab = "Time (days)",
  ylab = "Probability (PFS)",
  legend.title = "")

os_Surv_obj <- with(criteria_df,
                    survival::Surv(final_survival_time_os, final_censor_os))

os_survfit_obj <- survminer::surv_fit(os_Surv_obj ~ CGS_risk_C4,
                                      data = criteria_df)

os_plot <- survminer::ggsurvplot(
  fit = os_survfit_obj,
  data = criteria_df,
  conf.int = TRUE,
  pval = TRUE,
  surv.median.line = "hv",
  ggtheme = theme_bw(base_size = 16),
  risk.table = TRUE,
  title = "Overall Survival (OS)",
  xlab = "Time (days)",
  ylab = "Probability (OS)",
  legend.title = "")

combined_survival_plot_object <- (pfs_plot$plot + os_plot$plot) / free(pfs_plot$table + os_plot$table) +
  plot_layout(guides = "collect", heights = c(4,2)) +
  plot_annotation(title = "High risk due to Criteria 4 only vs. Standard risk") &
  theme_bw(base_size = 16) +
  theme(legend.position = "top")

ggsave(filename = here::here(plots_dir, str_c("criteria4_only_HR_SR_PFS_OS.pdf")),
       plot = combined_survival_plot_object,
       width = 12,
       height = 8)

combined_survival_plot_object

```

# Abstract numbers

```{r, warning = FALSE, message = FALSE}

n_hr <- risk_df |> filter(CGS_risk == "High") |> nrow()
n_sr <- risk_df |> filter(CGS_risk == "Standard") |> nrow()

n <- risk_df |> nrow()
n_na <- risk_df |> filter(is.na(CGS_risk)) |> nrow()

str_c("We evaluated the CGS risk status of ", n, " patients from CoMMpass, finding ",
      round(100*n_sr/(n_hr + n_sr), 1), "% (n = ", n_sr, ") to be standard risk and ",
      round(100*n_hr/(n_hr + n_sr), 1), "% (n = ", n_hr, ") to be high risk, after removing ",
      round(100*n_na/n, 1), "% (n = ", n_na, ") due to missing data.")

```

#### Median survival time (months)

```{r, warning = FALSE, message = FALSE}

summary(abstract_PFS_object)$table[,"median"]/30.4375

```

```{r, warning = FALSE, message = FALSE}

summary(abstract_OS_object)$table[,"median"]/30.4375

```


#### Criteria numbers

```{r, warning = FALSE, message = FALSE}


```
